No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-11-12 paper/GEB Submission/manuscript_GEB.Rmd
Language edits from editor

---
title: No general trend for multiple dimensions of functional diversity in bird and mammal communities
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing dissimilarity due to turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that underpin ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time.

@@

  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional change is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional changes independent of species richness, functional redundancy loss, and increasing species richness.

@@

To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species for abundance-based studies.
Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
`r colorize("This approach assumes the species pool is constant through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

While trends aggregating across studies did not indicate one prevailing kind of functional change, post hoc examination of trends for individual studies revealed four distinct change groups.
These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, functional change independent of species richness, loss of redundancy, and increase in species richness (Table \ref{tab:studyChange}).
The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
Studies in this group span the distribution of study durations, excluding only the very longest running studies, with the longest no-change time series lasting 33 years.
They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.

@@

The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
Based on a species approach alone, these communities appear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017].

@@

`r colorize("Study-level trends in CWM's offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover. Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM. This link between CWM's and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift. However, shifts in the functional space were not restricted to change studies, as we also found significant CWM trends in 7 of the 32 \"no change\" studies. Functional turnover was also qualitatively divorced from species turnover, with just as many studies exhibiting functional turnover (at least one significant CWM trend) with species turnover as not. These post hoc observations give evidence that even minimal changes in species composition undetected by trends in species-based metrics can translate to significant changes to one or a few trait axes.", "blue")`

@@

While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
For example, loss of evenness in communities with constant species richness may be a first sign of a community being impacted by environmental change, with negative implications for stability and function.

>>>>

---
title: No general trend for multiple dimensions of functional diversity in bird and mammal communities
abstract: |
  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing dissimilarity due to turnover but no overall trend in species diversity through time at the local scale. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that underpin ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional diversity and composition through time.

@@

  Main Conclusions: General temporal trends indicate that on the aggregate studies do not exhibit consistent changes in any aspect of functional diversity across many taxa, biomes, and realms. At the study-level, the majority of studies showed no temporal trends in species or functional diversity, with the remainder of the studies falling into broad categories of functional diversity change independent of species richness, loss of functional redundancy (species richness declines while functional richness is maintained), and increasing species richness sometimes accompanied by increasing functional richness.

@@

To assess changes in functional diversity independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species for abundance-based studies.
Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
`r colorize("This approach assumes the species pool is constant through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

While trends aggregating across studies did not indicate one prevailing kind of change in functional diversity or functional composition (with the exception of changes in fish consumption), post hoc examination of trends for individual studies revealed four distinct change groups.
These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, change in functional diversity independent of species richness, loss of redundancy (species richness declines while functional richness is maintained), and increase in species richness sometimes accompanied by increasing functional richness (Table \ref{tab:studyChange}).
The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
Studies in this group span the distribution of study durations, excluding only the very longest running studies, with the longest no-change time series lasting 33 years.
They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.

@@

The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
Based on a species approach alone, these communities appear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
However, with the addition of a functional perspective we found multiple kinds of underlying change in functional diversity including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017].

@@

`r colorize("Study-level trends in CWM's offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover. Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional diversity change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM. This link between CWM's and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift. However, shifts in the functional space were not restricted to change studies, as we also found significant CWM trends in 7 of the 32 \"no change\" studies. Functional turnover was also qualitatively divorced from species turnover, with just as many studies exhibiting functional turnover (at least one significant CWM trend) with species turnover as not. These post hoc observations give evidence that even minimal changes in species composition undetected by trends in species-based metrics can translate to significant changes to one or a few trait axes.", "blue")`

@@

While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for changes in functional diversity as a result of anthropogenic impact.
In fact, study-level trends indicate quite the opposite, that changes in functional diversity and composition with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
For example, loss of evenness in communities with constant species richness may be a first sign of a community being impacted by environmental change, with negative implications for stability and function.
